---
title       : Intro to R Workshop
subtitle    : UCI Data Science Initiative
author      : Sepehr Akhavan
job         : PhD Candidate, Dept. of Statistics
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : mathjax            # {mathjax, quiz, bootstrap}
logo     : logo.png
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
github:
  user: sepehrakhavan
  repo: IntroR_Workshop
---

## Introduction

```{r,echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(ggplot2)
```
1) The class will include 5 sessions:
  + Session 1 (9-10:20): Data Types in R 
  + Session 2 (10:30-12): Control Structures and Functions
  + Session 3 (12:30-1:50): Debugging Techniques in R
  + Session 4 (2-3:20): Plotting and Data Visualization in R 
  + Session 5 (3:30-5): Statistical Analysis in R

---

## Introduction

2) Each session will include:
  + Lecture: about 1 hr
  + R exercise and Discussion: 20 minutes
  + Break: 10 minutes


3) We are going to work in pairs. Please find a partner. 


4) Feel free to ask question anytime during lecture.

5) To access this presentation and the codes used during the workshop please visit:
  + https://github.com/SepehrAkhavan/intro-R-WorkShop
  + http://sepehrakhavan.github.io/intro-R-WorkShop/

---

## Session 1 - Agenda

1. Overview and History of R
2. Resources in Learning R and How to get Help?
3. RStudio
4. Data Types in R
5. Subsetting in R


---

## What is R?

+ R is a free software environment for statistical computing and graphics
  + See http://www.r-project.org/ for more info
  
  
+ R compiles and runs on a wide variety of UNIX platforms, Windows and Mac OS

+ R is Open-Source and free

+ R is fundamentally a command-driven system

+ R is an object-oriented programming language 
  + everything in R is considered as an object!


---

## History of R:

+ R created by Ross Ihaka and Robert Gentleman in 1991 in New Zealand
  + For more info please read: Ihaka, Ross, and Robert Gentleman. "R: a language for data analysis and graphics." Journal of computational and graphical statistics 5.3 (1996): 299-314. 

+ In 1996, the first R public mailing lists were created.
  + More info on R mailing lists: http://www.r-project.org/mail.html

+ In 1997, The R core team, with write access to the R source, was formed:
  + More info n the core team: http://www.r-project.org/contributors.html

+ In 2004, the first useR! conference was in Vienna, Austria.

+ In Feb 2011, the first version of RStudio was released.


---

## Online Resources to Learn R:

> 1. Very useful resources available on **The Comprehensive R Archive Network** (CRAN)
  + please visit: http://cran.us.r-project.org

> 2. R built-in facility:
  + try ?lm, help(lm) in R consol

> 3. There are many free tutorials available online:
  + Quick R: http://www.statmethods.net/
  + R-Twotorials: http://www.twotorials.com/
  + UCLA Academy Technology Services: http://www.ats.ucla.edu/stat/r/
  
> 4. R-Bloggers (http://www.r-bloggers.com/): is a central hub (e.g: A blog aggregator) of content collected from bloggers who write about R (in English). 


---

## Useful Books in learning R:

1. Chambers(1998). Programming with Data, Springer.

2. Venables & Ripley (2000). S Programming, Springer.

3. Chambers (2008). Software for Data Analysis, Springer. (highly recommended)

4. More resources available at: http://www.r-project.org/doc/bib/R-books.html

---

## How to get help in R:

1. Simply use the built-in help function in R
  + example: ?lm, help(lm)
  
2. R mailing lists: r-help and r-devel
  + For more info: https://stat.ethz.ch/mailman/listinfo/r-help
  + How to ask good questions: http://www.r-project.org/posting-guide.html

3. Use Q&A websites in particular:
  + stackoverflow (http://stackoverflow.com): for programming related questions.
  + crossValidated (http://stats.stackexchange.com): for Stats related questions.
  
4. Google :)  

---

## R Studio:

1. RStudio is a free and open source integrated development environment (IDE) for R.

2. To download RStudio please visit: http://rstudio.org/

3. Please note that you must have R already installed before installing R Studio.

---

## Data Types in R:

1. R has 5 main data types:
  + Numeric
  + Integer
  + Complex
  + Logical
  + Character
  
2. Everything in R is object. Objects can have some attributes.
  + names, dimension, class, length are some possible attributes

---

## Vectors in R:

Vector is the most basic object in R

```{r echo=TRUE}
numVec <- 1:10 # <- : is assigning operator
numVec

charVec <- c("a", "b", "c") # c: to combine elements
charVec

logVec <- vector(mode = "logical", length = 10)
logVec
```

---

### Special Values:

There are some special values in R:
  + use L to refer to an integer value: 1L
  + R knows infinity: Inf, -Inf
  + NaN: refers to "Not a number"
  
```{r echo=TRUE}
intVec <- c(1L, 2L, 3L, 4L) 
intVec

a <- Inf; b <- 0
rslt <- c(b/a, a/a)
rslt
```

---

### Logical, Complex, & Character Vectors:

Let's see some examples of logical, complex, and character vectors:
```{r echo=TRUE}
logVec <- c(TRUE, FALSE, FALSE, T, F)
logVec

compVec <- c(1 + 0i, 3 + 1i)
compVec

charVec <- c("red", "green", "blue")
charVec
```

---

### Data Type Coercion:

+ In general, vectors CAN NOT have mixed types of objects
+ exception: lists in R 

```{r echo=TRUE, results='hide'}
numCharVec <- c(3.14, "a")
numCharVec # ? what would you expect to be printed?

numLogVec <- c(pi, T)
numLogVec # any guess?

charLogVec <- c("a", TRUE)
charLogVec # ?
```

+ In examples above, we saw implicit coercion 
+ Explicit coercion is also possible!

---

### Data Type Coercion:

+ as(): To explicitly coerce objects from one type to another

```{r echo=TRUE}
numVec <- seq(from = 1200, to = 1300, by = 15)
numVec
numToChar <- as(numVec, "character")
numToChar
logVec <- c(F, T, F, T, T)
as(logVec, "numeric")
```

---

### Data Type Coercion:

+ Coercion does not always work! Be careful about warnings:

```{r echo=TRUE}
compVec <- c(12+10i, 1+6i, -3-2i)
as(compVec, "numeric")

charVec <- c("2.5", "3", "2.8", "1.5", "zero")
as(charVec, "numeric")
```

---

### Factors:

+ Factor is a vector object used to specify a discrete classification (categorical values).
+ Factors can be: 1) ordered, 2) un-ordered
+ Levels of a Factor are better to be labeled (self-descriptive)
  + Consider gender as (0, 1) as opposed to labeled ("F", "M")
```{r echo=TRUE}
Gender <- rep(c("Female", "Male"), times = 3)
Gender
GenderFac1 <- factor(Gender)
GenderFac1
```

---

### Factors:

```{r echo=TRUE}
levels(GenderFac1)
table(GenderFac1)
unclass(GenderFac1) # bring the factor down to integer values
```

---

### Factors:
```{r echo=TRUE}
GenderFac1 # levels are ordered alphabetically - 1st level = BaseLevel
GenderFac2 <- factor(Gender, levels = c("Male", "Female"))
GenderFac1
GenderFac2
```

---

### Missing Values:

+ There are two kinds of missing values in R:
  + NaN: refers to "Not a Number" and is a a missing value produced by numerical computation.
  + NA: When a value is "Not Available" or is "Missing", NA is assigned as its value.

+ NaN is also considered as NA (the reverse is NOT true). 
```{r echo=TRUE}
testScore <- NA
is.na(testScore)
is.nan(testScore)
```

---

### Matrices:

+ Matrix is a special case of vector:
  + Matrix has dimension attribute

```{r echo=TRUE}
myMat <- matrix(nrow = 2, ncol = 4)
myMat
attributes(myMat)
```

---

### Matrices:
```{r echo=TRUE}
myMat <- matrix(1:8, nrow = 2, ncol = 4)
myMat # matrices are filled in column-wise
```

---

### Matrix is a special vector:

```{r echo=TRUE}
myVec <- 1:8
myVec
dim(myVec) <- c(2,4)
myVec
```

+ Similar to vectors, all elements of a matrix should have the same type.
  + if not, R does an automatic coercion.

---

### Other Ways to Create Matrix:

+ As it's intuitive, matrix seems to be combination of vectors that are put next to each other (either column wise or row wise).

+ rbind() (row bind) and cbind (column bind) do a similar job:
```{r echo=TRUE}
vec1 <- 1:4
vec2 <- sample(1:100, 4, replace = FALSE)
vec3 <- rnorm(4, mean = 0, sd = 1)
colMat <- cbind(vec1, vec2, vec3)
colMat
```

---

### Other Ways to Create Matrix:

```{r echo=TRUE}
vec1 <- 1:4
vec2 <- sample(1:100, 4, replace = FALSE)
vec3 <- rnorm(4, mean = 0, sd = 1)
rowMat <- rbind(vec1, vec2, vec3)
rowMat
```

---

## Lists:

+ Consider list as a vector but with two main differences:
  + each element of a list can have its own class regardless of other elements
  + This means, each element can be of a different data type and a different length
```{r echo=TRUE}
myVec <- c(10, "R", 10-5i, T)
myList <- list(10, "R", 10-5i, T)
myVec
```

---

## Lists:

```{r echo=TRUE}
myList <- list(10, "R", 10-5i, T)
myList
```
+ Elements of list are shown with [[]]
+ Elements of vector are shown with []

---

## Data Frames:

+ We use data frames to store tabular data
+ Data frame is a special list where all objects have equal length
+ The main difference between data.frame and Matrix?
```{r echo=TRUE}
studentID <- paste("S#", sample(c(6473:7392), 10), sep = "")
score <- sample(c(0:100), 10)
gender <- sample(c("female", "male"), 10, replace = TRUE)
data <- data.frame(studentID = studentID, score = score, gender = gender)
head(data)
```

---

## Exercise 1:

+ a) Create two numeric vectors of the same length and find the product of these two vectors. 

+ b) Create two numeric vectors of DIFFERENT length and find the product of these two vectors. What do you expect to get? Can you explain the output that R generates?

+ c) Create two numeric matrices of 2 by 2 dimensions. 

+ d) Multiply the two matrices in part (c) using * operator

+ e) Multiply the two matrices in part (c) using %*% operator. What is the difference between part (d) and part (e)?

---

## Exercise 2:

+ a) please call help on sample() in R. What does this function do?
+ b) What does replace argument in sample() does?
+ c) Using sample(), randomly choose one number out of the pool of {1, 2, 3, 4, 5, 6}
+ d) Considering the experiment of rolling a die twice. How can you simulate it by using sample()? what is the role of replace argument in this case?

---

## Subsetting:

+ Often times we need to take a subset of a vector, a matrix, a list, or a dataframe.
+ We consider three main operators to take a subset of an object:
  + [ ]: single brackets return an object of the same class of the original object. By using [], we can also choose more than one element.
  + [[ ]]: double brackets are used primarily for lists and dataframes. 
  + $: Similar to double brackets, $ is used primarily for lists and dataframes. 

+ With [[ ]] and $, we can only choose one object!

+ [[ ]] and $ can return an object with a different class than the original objects we are subsetting from.

---

### Subsetting examples:

```{r echo=TRUE}
myVec <- 10:20
myVec[3]

myList <- list(obj1 = "a", obj2 = 10, obj3 = T, obj4 = 10-5i)
myList[[3]]
myList$obj3
```

---

## Subsetting with [ ]:

+ By using single bracket, we can choose more than one element of an object.
+ In this case, index vectors can be very useful:
  + Index vector is a vector of indices of another vector that is used to select a subset of another vector (or Matrix)
  
```{r echo=TRUE}
x <- seq(from=0, to=100,by=10) # length(x) is ??
IndVec <- c(1, 2, 3, 4, 5) # the first 5 elements 
x[IndVec]
```

---

## Index Vectors:

+ There are four types of Index vectors:
  1. Logical Index Vector: The logical index vector should be of the same length of the vector from which we are selecting a subset. Values corresponding to TRUE in the index vector are selected.
  2. Vector of Positive integers: All the values in this type of index vector must lie in 1:(length(x)).
  3. Vector of Negative integers: This type of index vector indicates the values to be excluded from the
vector.
  4. A Vector of Character Strings: if a vector has a name attribute, we can simply take a subset of the vector by calling the names of the elements.

---

## Index Vectors:
```{r echo=TRUE}
myVec <- letters[1:10]
names(myVec) <- paste("e", 1:10, sep = "")
myVec

logIndVec <- rep(c(T, F), each = 5)
logIndVec

posIndVec <- 1:5
negIndVec <- -6:-10
chIndVec <- c("e1", "e2", "e3", "e4", "e5")
```

---

## Index Vectors:
```{r echo=TRUE}
myVec[logIndVec]

myVec[negIndVec]

myVec[chIndVec]
```  

---

## Logical Index Vectors:

+ logical index vectors can be generated by using conditional statements:
  + Using ==, !=, <, >, ...

```{r echo=TRUE}
myVec <- 1:10
logIndVec <- (myVec < 5)
logIndVec
myVec[logIndVec]
```  

---

## Matrix Indexing:
+ Similar to vector indexing, we can refer to individual elements of a matrix.
```{r echo=TRUE}
myMat <- matrix(1:8, ncol = 4)
myMat
myMat[1,1] # refering to an element
myMat[2,] # refering to the second row
myMat[,3] # refering to the third column
```  

---

## Matrix Indexing:

+ By default, when the retrieved elements of a matrix look like a vector, R drops their dimension attribute. We can turn this feature off by setting drop = FALSE

```{r echo=TRUE}
myMat[1,1]
myMat[1,1, drop = FALSE]

myMat[2,, drop = FALSE]
``` 

---

## Subsetting Lists:
```{r echo=TRUE}
myList <- list(ch = letters[1:2], lg = F, nm = 1:3)
myList
myList[1] # subset is still a list
``` 

---

## Subsetting Lists:
```{r echo=TRUE}
myList[1:2] # subset is still a list
myList[[1]] # returning the 1st obj with its own class
myList$ch # alternative to [[]]
``` 

---

## Subsetting Lists:
```{r echo=TRUE}
myList[[1]][2] # returning the 2nd element of the 1st obj
myList$ch[2]
myList[[c(1,2)]]
``` 

---

## Subsetting Data Frames:
```{r echo=TRUE}
library(datasets)
data(quakes) # ?quakes for more info
str(quakes)
head(quakes$long)
``` 

---

## Subsetting Data Frames:

```{r echo=TRUE}
library(datasets)
quakes[1:10,]
``` 

---

## Exercise 3:

+ a) Please run the commands below to load in titanic3 dataset
```{r echo = TRUE, warning=FALSE, error=FALSE, message=FALSE, eval=FALSE}
install.packages("PASWR", repos = "http://cran.stat.ucla.edu") # installs PASWR package
library(PASWR) # this loads in the package
```

+ b) print out the top 10 rows of the dataset.

+ c) print out the names of the columns of the dataset
  + hint: use names()

+ d) print the top 10 values of age column

+ e) using is.na() function, take a subset of the dataset that has non-missing age
  + hint: !is.na() refers to NOT NA !

+ e) Run ?which to use help on understanding what which() does.


---

## Time to Break for 10 Minutes :)

---

## Session 2 - Agenda

1. Vectorized Operations in R
2. Reading and Writing in R
3. Control Structure
4. Functions & Packages

---

## Vectorized Operations

R is capable of vectorized operations without any need for running loops:
```{r echo=TRUE}
x <- 1:5
y <- c(1, 2, 6, 7, 10)
x + y # R does an element by element summation
x < y
```

---

## Vectorized Operations

+ Similar to vectors, vectorized operations can be performed for Matrices:
```{r echo=TRUE}
x <- matrix(1:9, ncol = 3)
y <- matrix(rep(c(5,6,7), 3), ncol = 3)
x + y # R does an element by element summation
x < y
```

---

## Reading and Writing Data

**The slides for "Reading and Writing Data" section were mainly from Dr. Roger D. Peng, Associate Professor at Johns Hopkins**

Main functions on reading data into R:
1. read.table(), read.csv(): to read tabular data 
2. readLines(): to read lines of a text file
3. source(), dget(): reading R codes
4. load(): to read saved workspaces

+ Only read.table() and read.csv() are covered in this lecture. 

---

## Reading and Writing Data

Main functions on writing data from R:

1. write.table(), write.csv(): to write tabular data to file
2. writeLines(): to write lines to a text file
3. dump, dput: to write R codes to a file
4. save: to save a workspace

+ Only write.table() is covered in this lecture. 

---

## read.table():

+ read.table() is the most commonly used function to read data in R. Below are important arguments of this function:

  + file: name or address to the file of interest
  + header: logical indicator on whether the file has header or not
  + sep: string on how columns of data are separated (in .csv, sep = ",")
  + colClasses: is a character vector for class of each column
  + nrows: number of rows in the dataset
  + comment.char: a character that is used in the dataset for commenting
  + skip: number of lines to skip from the beginning of the file
  + stringAsFactors: logical indicator on whether characters should be converted to factors 

+ read.csv() is equivalent to read.table with sep = ","

---

## read.table():

```{r echo = FALSE}
irisFile <- read.table(file = "/Users/sepehrakhavan/Documents/Statistics/Stat Courses/DataScience_Dr.PadhraicSmyth/RWorkShop_Nov_2014/SupportingFiles/iris", header = TRUE)

head(irisFile)
```

+ to make read.table() run faster:
  + set comment.char = ""
  + set colClasses upfront

---

## Calculating Memory Requirements:

+ Note that datasets will be read into RAM. So, you should have enough RAM in order to read a dataset. 

+ Consider a data frame with 1.5 million rows and 120 columns. How much memory is required to read this dataset?

1.5m * 120 * 8 bytes/numeric = 1.44 * $10^9$ = 1.44 * $10^9$/ $2^{20}$ MB = 1,373.29 MB = 1.34 GB

+ So it's recommended to have a RAM of size 2 * 1.34GB to read that dataset.

---

## write.table():

```{r echo = T, eval=FALSE}
write.table(irisFile, file = "pass/to/the/file")
```

---

## Loops:

+ There are 3 ways in R to write loops:
  + repeat
  + while
  + for

---

### repeat:

+ The repeat statement causes repeated evaluation of the body until a break is specifically requested.
+ The body of the repeat statement should include:
  + statements to do computation
  + conditions on when to stop

```{r echo = T}
i <- 1 
repeat {
  print(paste("cycle #", i, sep = "")) 
  i=i+1
  if (i >= 2)
  break 
}
```

---

### while:

+ while loops start by testing a condition. If the condition is true, while executes the body:

```{r echo = T}
i <- 1
while (i < 2) {
  print(paste("cycle #", i, sep = ""))
  i <- i + 1 
}
```

---

### for:

```{r echo = T}
for(i in 1:4){
  print(paste("cycle #", i, sep = ""))
  i <- i + 1 
}
```

---

## next:

+ next is useful in skipping an iteration of a loop:
```{r echo = T}
for(i in 1:4){
  if (i == 3){
    next
  }
  print(paste("cycle #", i, sep = ""))
  i <- i + 1 
}
```


---

## if:

+ if/else statements are used to write conditional statements

```{r echo = T}
x <- 7
if (x < 10){
  print("x is less than 10")
}else{
  print("x is greater than 10")
}
```

---

## if:
```{r echo = T}
age <- sample(1:100, 10)
ageCat <- rep(NA, length(age))
for (i in 1:(length(age))) {
    if (age[i] <= 35){
       ageCat[i] <- "Young"
      }else if (age[i] <= 55){
        ageCat[i] <- "Middle-Aged"
      }else{
         ageCat[i] <- "Old"
      } 
}
age.df <- data.frame(age = age, ageCat = ageCat)
age.df[1:3,]
```

---

## Functions and Packages:

> 1. R language has many built-in functions
> 2. Each function has a name followed by ()
> 3. Arguments of a function should mention within parentheses
> 4. R packages are a comfortable way to maintain collections of R functions and data sets
> 5. Packages allow for easy, transparent and cross-platform extension of the R base system


---

## Functions and Packages:

There are some terms which sometimes get confused and should be clarified:

1. Package: An extension of the R base system with code, data and documentation in a standardized format
2. Library: A directory containing installed packages
3. Repository: A website providing packages for installation
4. Source: The original version of a package with human-readable text and code
5. Binary: A compiled version of a package with computer-readable text and code, may work only on a specific platform
6. Base packages: Part of the R source tree, maintained by R Core

+ for more info on how R packages are developed, please read: "Creating R Packages: A Tutorial" (Friedrich Leisch)
  + http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf


---

## How to install a package in R:

There are two ways to install a package in R:

1. Installing from CRAN: install a package directly from the repository
  + Using R studio: tools/install packages
  + From R console: install.packages()

2. Installing from Source: In this method, you should first download the add-on R package and use the following unix command in the console to install the package:
  + R CMD INSTALL packageName -l path/to/your/Rpackage/Directory

+ Once you install a package, you need to load it into R using the function library()


---

## Exercise 4:

+ Please install ggplot2 package
+ Once installed, please load the package


---

## Calling a function in R

```{r echo=TRUE}
str(sample)
```
+ consider sample() in R. Simply run ?sample in R console to read the help on this function.
+ sample() gets four arguments: 
  + x: sample space in form of a vector
  + size: your desired sample size
  + replace: sampling with/without replacement
  + prob: a vector of probability weights

+ some of the arguments have default values. What are those arguments?

+ How to use (or call) this function? 


---

## Calling a function in R

```{r echo=TRUE}
# Functions arguments can be matched: 1) by position or 2) by name
sampSpace <- 1:6 # rolling a die
sample(sampSpace, 1) # arguments with default values can be omitted
sample(size = 1, x = sampSpace) # no need to remember the order 
sample(size = 1, sampSpace)
```


---

## Writing your Own functions

```{r echo=TRUE, eval=FALSE}
yourFnName <- function(<your arguments>){
  # body of your code
  
  # return the output of the function
}

# to use your function, you can simply call the function name as:
yourFnName(<set values for the input arguments>)
```


---

## Writing your Own functions

+ Let's write a function that gets three arguments: a, b, c
+ The function then returns min of these two numbers

```{r echo=TRUE, eval=TRUE}
myMin <- function(a, b, c){
  myMinVal <- min(a, b, c)
  return(myMinVal)
}

myMin(10, 20, 30)
myMin(10, NA, 20) # ? how to fix this so it returns 10
```


---

## Writing a function for 2-sample t-test
```{r echo=TRUE, eval=TRUE}
myTtest <- function(vec1, vec2) {
    # my T-Test function / Assuming constant variance + Two-Sided T Test
    n1 <- length(vec1)
    n2 <- length(vec2)
    miu1 <- mean(vec1)
    miu2 <- mean(vec2)
    var1 <- var(vec1)
    var2 <- var(vec2)
    pooledVar <- ((n1 - 1) * (var1) + (n2 - 1) * (var2))/(n1 + n2 - 2)
    testStat <- (miu1 - miu2)/(sqrt(pooledVar * (1/n1 + 1/n2)))
    df <- n1 + n2 - 2
    pval <- 2 * pt(abs(testStat), df, lower.tail = FALSE)
    rslt <- c(`Mean of Vec1` = miu1, `Mean of Vec2` = miu2, testStat = testStat,
        df = df, pval = pval)
    return(rslt)
}
```


---

## Writing a function for 2-sample t-test
```{r echo=TRUE, eval=TRUE}
y1 <- rnorm(10, mean = 0, sd = 1); y2 <- rnorm(10, mean = 0, sd = 1)
myTtest(y1, y2)

# Using R's built-in T-Test:
# t.test(y1, y2, var.equal = TRUE)  # ? Test this on your own!
```


---

## Some notes on Functions

> 1. Variables defined within a function are locally defined (i.e. not defined outside of the function).

> 2. Functions in R are treated like any other first class objects. This means functions can be an passed as arguments of other functions.

> 3. Arguments of functions are evaluated as they are needed (lazy evaluation). 

> 4. " ... " can be an argument of a function and it refers to a situation where number of input arguments can be varied and is not fixed upfront. 


---

## Lazy Evaluation of Function Arguments

```{r echo=TRUE}
myLazyFn1 <- function(a, b){
  return(a)
}
myLazyFn1(10) # No error!

myLazyFn2 <- function(a, b){
  print(a)
  print(b)
  return(1)
}
myLazyFn2(10) # No error!
```


---

## Some useful functions:

+ Here we are going to talk about:
  + str(): a function to explain internal structure of a function
  + apply(): to apply a function to a matrix or dataframe
  + lapply(), sapply(), tapply(), mapply(): applying a function to a vector
  + split(): to split a dataset by levels of a factor


---

### str():

+ str() is a compact way of understanding what an object is and what is in that object.
```{r echo=TRUE}
str(str)
str(sample)

genderF <- factor(sample(c("Male", "Female"), 20, replace = TRUE))
str(genderF)
```


---

### str():
```{r echo=TRUE}
myMat <- matrix(1:10, ncol = 5)
str(myMat)
myList <- list(numVec = 1:3, logVec = F, charVec = LETTERS[1:4])
str(myList)
```


---

### apply():

```{r echo=TRUE}
str(apply) # try ?apply for more info
```

+ apply() is a useful function to apply a function (FUN) on a MARGIN of a matrix or dataframe (X)

+ MARGIN: a vector giving the subscripts which the function will be applied over
  + 1: indicates rows
  + 2: indicates columns
  + c(1, 2): indicates rows and columns

+ FUN: refers to the function that we want to apply on the dataset

+ "..." : additional arguments of FUN


---

### apply():

```{r echo=TRUE}
myMat <- matrix(1:10, ncol = 5)
myMat[2,c(2, 5)] <- NA
myMat
apply(myMat, 2, sum, na.rm = TRUE)
```


---

### apply():

```{r echo=TRUE}
# consider iris dataset: 
head(iris) # more info ?iris
# suppose we are interested in getting 25% and 75% of each numeric column
```


---

### apply():

```{r echo=TRUE}
# Consider iris dataset: 
apply(iris[,-5], 2, quantile, probs = c(0.25, 0.75))
```


---

### lapply() and sapply():

```{r echo=TRUE}
str(lapply)
str(sapply)
```
+ x: a list, dataframe, or a vector
+ FUN: the function to be applied to each element of X
+ "...": other arguments of FUN


---

### lapply() and sapply():
```{r echo=TRUE}
myList <- list(e1 = 1:10, e2 = -1:-10)
lapply(myList, mean)
sapply(myList, mean)
```


---

### lapply() v. sapply()?:

+ sapply() simplfies the result of lapply.
  
+ If the result of lapply is a list with all elements of the same length:
    + if length == 1: sapply() returns a vector
    + if length != 1: sapply() returns a matrix
    
+ otherwise, sapply() generates a list similar to lapply()


---

### lapply() & sapply() with a user-defined FUN
```{r echo=TRUE}
myList <- list(e1 = 1:10, e2 = -1:-10)
lapply(myList, function(element){return(mean(c(element[1], element[length(element)])))})
sapply(myList, function(element){return(mean(c(element[1], element[length(element)])))})
```


---

### tapply():
```{r echo = TRUE}
str(tapply)
```

+ tapply() applies a function on a subset of of a vector
+ X: is a vector 
+ INDEX: list of one or more factors, each of same length as X
+ FUN: our function of interest
+ "...": other arguments of FUN
+ simplify: any guess???


---

### tapply():
```{r echo = TRUE}
HeightDF <- data.frame(heigth = c(rnorm(100, 180, 3), rnorm(100, 170, 3)), gender = factor(rep(c("M", "F"), each = 100)))
head(HeightDF)
tapply(HeightDF$heigth, HeightDF$gender, mean)
```


---

### mapply():
```{r echo=TRUE}
str(mapply)
```
  
+ all previous "apply" functions were univariate
  + f(x, {some other parameters})
  
+ What to do if we want to apply a multivariate function:
  + f(x, y, {some other parameters}) # we can have more than 2 variables 


---

### mapply():
```{r echo=TRUE}
l1 <- list(e1 = 1:10, e2 = 1:10)
l2 <- list(e1 = -1:-10, e2 = -1:-10)
# how to get l1$e1[i] + l1$e2[i] + l2$e1[i] + l2$e2[i] ? 
mapply(sum, l1$e1, l1$e2, l2$e1, l2$e2)
```


---

### split():
```{r echo=TRUE}
str(split) # ?split for more info
```

+ X: a vector or a data frame
+ f: factor
+ drop: should R drop empty factor levels?


---

### split():
```{r echo=TRUE}
str(HeightDF)
# Goal: to separate Females from Males
splitData <- split(HeightDF$heigth, HeightDF$gender)
str(splitData)
```


---

## Time for Lunch Break for 30 Minutes. Please come back at 12:30 :)

---

## Session 3 - Agenda

1. Useful Matrix Functions
2. Statistical Distributions in R
3. Exercise

---

## Useful Matrix Functions

Consider A as a matrix. We can then have:
> 1. t(A): transpose of A
> 2. solve(): to get inverse of A
> 3. eigen(): to get eigen values and eigen vectors (if A is symmetric)
> 4. We only cover solve() in this lecture

---

### solve():
+ Consider B = A %*% X (where X is an unknown matrix)
+ Then: X = solve(A, B)
+ In a special case where B = I, X = $A^{-1}$

```{r echo=TRUE}
A <- matrix(c(1, 2, 3, 2, 4, 5, 3, 5, 6), ncol = 3)
A
# to get inverse of A: solve(A)
```

---

### solve():
```{r echo=TRUE}
solve(A)
# To check that solve(A) is inverse of A:
solve(A)%*%A
```
+ Machine epsilon is defined to be the smallest positive number which, when added to 1, gives a number different from 1.
+ Please visit http://en.wikipedia.org/wiki/Machine_epsilon for more info

---

## Statistical Distributions in R:

+ R has many in-built statistical distributions
  + examples: binomial, poisson, normal, chi square, ...

+ Each distribution in R has four functions:
  + these functions beging with a "d", "p", "q", or "r" and followed by the name of the distribution

+ ddist(parameters): refers to the density of each distribution
+ rdist(parameters): generates random numbers out of each distribution
+ qdist(parameters): to get quantile of a distribution
+ pdist(parameters): to calculate CDF


---

### Example of a Discrete Distribution:
```{r echo=TRUE}  
# Consider tossing a coin 10 times
str(dbinom)
dbinom(5, 10, 0.5) # prob of getting two head

str(pbinom) # cumulative dist
pbinom(5, 10, 0.5) # Pr[X <= 5]
```


---

### Example of a Discrete Distribution:
```{r echo=TRUE}  
str(qbinom) # quantile: Pr[X <= ?] = known value
qbinom(0.6230, 10, 0.5) # prob of getting two head

str(rbinom) # Generating random numbers
rbinom(20, 10, 0.5) # Pr[X <= 5]
```


---

### Example of a Continuous Distribution:
```{r echo=TRUE}  
# Consider a standard Normal distribution
str(dnorm)
dnorm(x = 0, mean = 0, sd = 1, log = FALSE)

str(pnorm) # cumulative dist
pnorm(0, mean = 0, sd = 1)
```


---

### Example of a Continuous Distribution:
```{r echo=TRUE}  
str(qnorm) # quantile
qnorm(0.5, mean = 0, sd = 1)

str(pnorm) # cumulative dist
rnorm(10, mean = 0, sd = 1)
```


---

### Example of a Continuous Distribution:
```{r echo=TRUE, fig.height = 4.5, fig.align='center'}  
# Let's try plotting Normal curve (more on plotting later)
x <- seq(from = -3, to = 3, by = 0.05)
y <- dnorm(x, mean = 0, sd = 1)
plot(x, y, type = "l")
```


---

## Exercise:
### Source: Dr. Stacey Hancock, Dept. Statistics - UCI
+ Suppose 20 people go to a fancy restaurant. Each person is wearing a hat and checks his/her hat at the door as he/she arrives. The hat-check attendant gets tipsy throughout the evening, forgetting which hat belongs to whom, and returns a random hat to each person as they leave. The patrons leave in a random order.

+ Conduct a simulation in R to answer the following question:
  + estimate the probability that at least one patron gets his or her own hat back
  + calculate the empirical mean (average) number of patrons who get their own hats back.
  


---

## Exercise 
### Source: Dr. Stacey Hancock, Dept. Statistics - UCI
+  Write an R function that will simulate rolling two k-sided dice (each numbered 1, 2, 3,...., k) N times, and return the proportion of times your sum was greater than or equal to 7. 
+ Set the function's default value for k to 6 and the default value for N to 1000. 


---

## Time for Lunch Break for 10 Minutes. Please come back at 2:30 :)

---

## Session 4 - Agenda

1. Plotting Systems in R
2. Base Plotting System in detail
3. Graphic Devices in R
4. Examples of other plotting Systems

---

## Plotting Systems in R

There are three main plotting systems in R:
  + Base Plotting System
  + Lattice
  + ggplot2

Plotting and Graphics engines in R is in the following packages:
  + graphics: contains plotting functions for the Base system
  + grDevices: contains graphic devices in R
  + lattice: contains plotting functions of the lattice system
  + grid: lattice package was built on top of this package
  + ggplot2

---

### Base System (intro)

In Base plotting system:
  + We start with a blank canvas
  + by calling a plot function, we start our plot
  + Using annotation functions we can add more elements to our plot
  
In general, plotting functions under the base system are:
  + Functions to generate plots: plot()
  + Functions to annotate plots: text(), lines(), points()


---

### Base System (example)

```{r echo=TRUE, fig.keep='last', fig.height=4.5, fig.align='center'}
x <- rnorm(100)
y <- rnorm(100)
plot(x, y)
abline(h = 0)
```


---

### Lattice (intro)

+ Lattice plotting system is part of the lattice package
+ Unlike the base system, each plot is constructed with a single function call
+ Plots can not be modified once they are created
+ Since we only have one function call, we usually need to specify a lot of parameters in the plotting function call under this package


---

### ggplot2 (intro)

+ ggplot2 is a very popular graphics system in R (written by Hadley Wickham)
+ The package has a language or grammar for describing different aspects of plotting
+ The package is an implementation of the Grammar of Graphics by Leland Wilkinson
+ ggplot2 mixes ideas from Base and from Lattice plotting systems
  + Many of the aesthetics of a plot (labeling, margins, background, ...) are done automatically with a single function call (similar to lattice)
  + You can still add components of a plot layer by layer (similar to the base system)


---

## Base Plotting System - Histogram

```{r echo=TRUE, fig.keep='last', fig.height=4.5, fig.align='center'}
x <- rnorm(100)
hist(x)
```


---

## Base Plotting System - Scatter Plot
```{r echo=TRUE, fig.keep='last', fig.height=4.5, fig.align='center'}
x <- rnorm(100)
y <- rnorm(100)
plot(x, y)
```


---

## Base Plotting System - BoxPlot
```{r echo=TRUE, fig.keep='last', fig.height=4.5, fig.align='center'}
x <- rnorm(100, 0, 1)
y <- rnorm(100, 1, 1)
z <- rnorm(100, 2, 2)
boxplot(x, y , z)
```


---

## Base Plotting System - Important Parameters

+ pch: plotting symbol (pch = 1 for open circle)
+ lty: line type (solid, dashed, dotted, ...)
+ lwd: line width (takes an integer number)
+ col: plotting color 
  + run colors() to get a list of R colors
+ type: type of plot
  + "p": points
  + "l": lines
  + "b": both points and lines
  + "n": no plotting
  

---

## Base Plotting System - Important Parameters

+ par(): to specify global graphics parameters that affect all plots in an R session
+ you can run par() to get a list of current values of these paraemters
+ run ?par for more information from help
+ Some important parameters are: 
  + mar: size of margins of a plot
  + mfrow: multiple plots that are arranged row-wise
  + mfcol: multiple plots that are arranged column-wise
  + bg: the background color
  

---

## Base Plotting System - Important Parameters
Let's see the default values for the parameters introduced so far:
```{r echo=TRUE}
t(par(c("pch", "lty", "lwd", "col")))
par("mar")
par("mfrow")
```
  

---

## Base Plotting System - Important Parameters

```{r echo=TRUE}
par("mfcol")
par("bg")
```

  

---

## Base Plotting System - Important Functions

+ plot: usually to make scatter plot
  + depending on the type of it input, might behave differently

+ lines: to add lines to a plot (ex: to overlay a density on a histogram)

+ points: to add points to a plot

+ text: to add texts to a plot at (x, y)

+ title: to add title to a plot

+ mtext: to add text to a margin of a plot

+ axis: to add customized axis to a plot

  

---

## Base Plotting System - Demo



  

---

## Graphics Devices in R

+ Graphics devices are places where plots in R appear:
  + screen device
  + file device

+ Screen Devices:
  + quartz() (Mac OS), x11() (Unix/Linux), windows() (Windows)

+ File Devices:
  + pdf(), png(), svg(), ...
  
+ By default, plots are sent to the screen device
  
+ for more info, please see ?Devices

---

## Graphics Devices in R

+ There is nothing special to do to use the screen device
+ For file devices:
  + Graphics device of interest should be opened
  + Plotting codes
  + closing the device using dev.off()

```{r echo=TRUE, eval=FALSE}
x <- rnorm(100)
pdf(file = "path/to/file.pdf")
hist(x)
dev.off()
```

---

## File Devices

+ File devices are in two main types:
  + Vector formats: pdf(), svg(), postscript()
  + Bitmap formats: png(), jpeg(), bmp()


---

## Lattice Plotting System
+ In Lattice plotting system, plotting is a one-step process with a single function call
+ Important functions:
  + xyplot(): to create scatterplot
  + bwplot(): to create boxplot
  + histogram()
  + stripplot(): boxplot with overlaid data points


---

### Lattice Scatter plot
```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height=5}
library(lattice)
data(diamonds)
diamonds_subset <- diamonds[sample(1:nrow(diamonds),500),]
xyplot(price ~ carat, data = diamonds_subset)
```


---

### Lattice Scatterplot across 
```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height = 5,fig.width = 12}
xyplot(price ~ carat | cut, data = diamonds_subset, layout = c(5,1))
```


---

### Lattice Scatterplot with Linear fit
```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height = 5,fig.width = 12}
xyplot(price ~ carat | cut, panel = function(x, y, ...){
  panel.xyplot(x, y, ...) # Here we call the default panel function
  panel.lmline(x, y, col = "red") # We overlay lm fit
}, data = diamonds_subset, layout = c(5,1))
```


---

## ggplot2
+ As Hadley Wickham explained in his ggplot2 book: ggplot2 is a package to "shorten the distance from mind to page"
+ Here we briefly mention some examples. For more information please visit: http://ggplot2.org
+ We focus on qplot() function to:
  + create scatter plot
  + create histogram


---

### qplot()
+ qplot() stands for quick plot
+ it's analogous to plot() in the base package
+ main elements of each plot are:
  + aesthetics: refers to size, shape, color, ...
  + geoms: refers to points, lines, bars, ...
  + for a complete list of aesthetics and geoms as well as other elements, please visit: http://docs.ggplot2.org/current/  
  

---

### qplot() - Scatterplot
```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height=5}
library(ggplot2)
qplot(carat, price, data = diamonds_subset)
```
  

---

### qplot() - Scatterplot
```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height=5}
qplot(carat, price, data = diamonds_subset, col = cut)
```
  

---

### qplot() - Facets
```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height=5, fig.width=12}
qplot(carat, price, data = diamonds_subset, facets = . ~ cut) # rows ~ columns
```

  

---

### qplot() - Histogram
```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height=5}
qplot(price, data = diamonds_subset, fill = cut, binwidth = 1000)
```


---

## Time for Lunch Break for 10 Minutes. Please come back at 4 :)


---

## Session 5 - Agenda

1. T-Test in R
2. ANOVA in R
3. Linear Regression in R


---

## T-Test in R

T-tests can be categorized into two groups:
  + 1) One-Sample t-test
  + 2) two-sample t-test


---

###  One-Sample T-Test
```{r echo=TRUE}
oneSampData <- rnorm(100, mean = 0, sd = 1)
oneSampTest.0 <- t.test(oneSampData) # ?t.test
oneSampTest.0
```


---

###  One-Sample T-Test
```{r echo=TRUE}
names(oneSampTest.0) # alternative to names()?? 
```  


---

###  Two-Sample T-Test
Two sample t-tests are categorized into 3 groups:
  + T-Test with equal variances
  + T-Test with un-equal variances
  + Paired T-Test: can be also considered as one-sample t-test on deltas.


---

###  Two-Sample T-Test (Un-equal Variances)
```{r echo = TRUE}
Samp1 <- rnorm(30, mean = 2.5, sd = 1)
Samp2 <- rnorm(50, mean = 5.5, sd = 1)
t.test(Samp1, Samp2)  # default assump: unequal variances
```


---

###  Two-Sample T-Test (Equal Variances)
```{r echo = TRUE}
t.test(Samp1, Samp2, var.equal = TRUE)  # default assump: unequal variances
```


---

###  Two-Sample T-Test (Paired T Test)
```{r echo = TRUE}
t.test(Samp1, Samp2[1:30], paired = TRUE)  # default assumption: unequal variances
```


---

##  ANOVA
If you are not familiar with ANOVA, simply consider ANOVA as an extension to two-sample t-test where we have more than two groups.

```{r echo = TRUE}
Samp1 <- round(rnorm(10, mean = 25, sd = 1), 1)
Samp2 <- round(rnorm(10, mean = 30, sd = 1), 1)
Samp3 <- round(rnorm(10, mean = 35, sd = 1), 1)
myDF <- data.frame(y = c(Samp1, Samp2, Samp3), group = rep(c(1, 2, 3), each = 10))
myDF$group <- as.factor(myDF$group)
str(myDF)
```


---

##  ANOVA
```{r echo = TRUE}
ANOVAfit <- lm(y ~ group, data = myDF)  # instead of lm, aov() can also be used!
myANOVA <- anova(ANOVAfit)  # anova computes analysis of variance tables on a fitted model object.
str(myANOVA) # see what is 
```


---

##  ANOVA
+ To learn more on how to fit ANOVA, please visit: 
  + http://www.statmethods.net/stats/anova.html
  

---

##  Linear Regression:
+ lm() is used to fit linear regression
+ Here we use "Prestige" dataset from "car" package
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
# install.package("car")
library(car)
data(Prestige) # load the data
str(Prestige)
```
  

---

###  Linear Regression - Fit:
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
myReg <- lm(prestige ~ education + income + women, data = Prestige)
myReg # summary(myReg)
names(myReg)
```
  

---

###  Linear Regression - Fit:
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
head(myReg$fitted.values)
head(myReg$residuals) # y - y.hat
```
  

---

##  Linear Regression - Diagnostics:
+ Here we cover some common regression diagnostics including:
  + Testing for Normality
  + Testing for Constant Variance
  
+ Reference: http://www.statmethods.net/stats/rdiagnostics.html
  

---

###  Testing for Normality

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE, fig.height=5}
qqPlot(myReg, main = "QQ Plot")
```
  

---

###  Testing for Constant Variance

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE, fig.height=5, warning=FALSE, message=FALSE, error=FALSE, results='hide'}
spreadLevelPlot(myReg)
```
