---
title: 'Intro to R Workshop: Session 3'
output: slidy_presentation
date: "May 30, 2016"
subtitle: UCI Data Science Initiative
---

```{r, include=FALSE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

---

## Session 2 - Agenda

1. Vectorized Operations in R
2. Reading and Writing in R
3. Control Structure
4. R Packages and Functions

---

## Vectorized Operations

R is capable of vectorized operations without any need for running loops:
```{r echo=TRUE}
x <- 1:5
y <- c(1, 2, 6, 7, 10)
x + y # R does an element by element summation
x < y
```

---

## Vectorized Operations

+ Similar to vectors, vectorized operations can be performed for Matrices:
```{r echo=TRUE}
x <- matrix(1:9, ncol = 3)
y <- matrix(rep(c(5,6,7), 3), ncol = 3)
x + y # R does an element by element summation
x < y
```

---

## Reading and Writing Data

**The slides for "Reading and Writing Data" section were mainly from Dr. Roger D. Peng, Associate Professor at Johns Hopkins**

Main functions for reading data into R:

1. read.table(), read.csv(): to read tabular data 
2. readLines(): to read lines of a text file
3. source(), dget(): reading R codes
4. load(): to read saved workspaces

+ Only read.table() and read.csv() are covered in this lecture. 

---

## Reading and Writing Data

Main functions for writing data from R:

1. write.table(), write.csv(): to write tabular data to file
2. writeLines(): to write lines to a text file
3. dump, dput: to write R codes to a file
4. save: to save a workspace

+ Only write.table() is covered in this lecture. 

---

## read.table():

+ read.table() is the most commonly used function to read data in R. Below are important arguments of this function:

  + file: name or address to the file of interest
  + header: logical indicator on whether the file has header or not
  + sep: string on how columns of data are separated (in .csv, sep = ",")
  + colClasses: is a character vector for class of each column
  + nrows: number of rows in the dataset
  + comment.char: a character that is used in the dataset for commenting
  + skip: number of lines to skip from the beginning of the file
  + stringAsFactors: logical indicator on whether characters should be converted to factors 

+ read.csv() is equivalent to read.table with sep = "," and header = TRUE

---

## read.table():

```{r echo = TRUE, eval=TRUE}
irisFile <- read.table(file = "iris.csv", sep=",", header = TRUE)
head(irisFile)
```

+ to make read.table() run faster:
  + set comment.char = " "
  + set colClasses upfront

---

## Calculating Memory Requirements:

+ Note that datasets will be read into RAM. So, you should have enough RAM in order to read a dataset. 

+ Consider a data frame with 1.5 million rows and 120 columns. How much memory is required to read this dataset?

1.5m * 120 * 8 bytes/numeric = 1.44 * $10^9$ = 1.44 * $10^9$/ $2^{20}$ MB = 1,373.29 MB = 1.34 GB

+ So it's recommended to have a RAM of size 2 * 1.34GB to read that dataset.

---

## write.table():

```{r echo = T, eval=FALSE}
write.table(irisFile, file = "path/to/the/file")
```

---

## Loops:

+ There are 3 ways in R to write loops:
  + for 
  + repeat (skipped!)
  + while (skipped!)

---

### for:

```{r echo = T}
for(i in 1:4){
  print(paste("cycle #", i, sep = ""))
  i <- i + 1 
}
```

---

## if:

+ if/else statements are used to write conditional statements

```{r echo = T}
x <- 7
if (x < 10){
  print("x is less than 10")
}else{
  print("x is greater than 10")
}
```

---

## if:
```{r echo = T}
age <- sample(1:100, 10)
ageCat <- rep(NA, length(age))
for (i in 1:(length(age))) {
    if (age[i] <= 35){
       ageCat[i] <- "Young"
      }else if (age[i] <= 55){
        ageCat[i] <- "Middle-Aged"
      }else{
         ageCat[i] <- "Old"
      } 
}
age.df <- data.frame(age = age, ageCat = ageCat)
age.df[1:3,]
```

---

## Functions and Packages:

1. R language has many built-in functions
2. Each function has a name followed by ()
3. Arguments of a function are put within parentheses
4. R packages are a comfortable way to maintain collections of R functions and data sets
5. Packages allow for easy, transparent and cross-platform extension of the R base system


---

## Functions and Packages:

There are some terms which sometimes get confused and should be clarified:

1. Package: An extension of the R base system with code, data and documentation in a standardized format
2. Library: A directory containing installed packages
3. Repository: A website providing packages for installation
4. Source: The original version of a package with human-readable text and code
5. Base packages: Part of the R source tree, maintained by R Core

+ for more info on how R packages are developed, please read: "Creating R Packages: A Tutorial" (Friedrich Leisch)
  + http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf


---

## How to install a package in R:

There are three main ways to install a package in R:

1. Installing from CRAN: install a package directly from the repository
  + Using R studio: tools/install packages
  + From R console: install.packages()

2. Installing from Source: In this method, you should first download the add-on R package and use the following unix command in the console to install the package:
  + R CMD INSTALL packageName -l path/to/your/Rpackage/Directory

3. Installing from a version control (Github): 
  + Check-out https://github.com/hadley/devtools

+ Once you install a package, you need to load it into R using the function library()


---

## Popular Packages in R:

1. To visualize data:
  + ggplot2: to create beautiful graphics
  + googleVis: to use Google Chart tools
  
2. To report results:
  + shiny: to create interactive web-based apps
  + knitr: to combine R codes and Latex/Markdown codes
  + slidify: to build HTML 5 slide shows
  
3. To write high-performance R code:
  + Rcpp: to write R functions that call C++ code
  + data.table: to organize datasets for fast operations
  + parallel: to use parallel processing in R
  

---


## Calling a function in R


```{r echo=TRUE}
str(sample)
```

+ consider sample() in R. Simply run ?sample in R console to read the help on this function.
+ sample() gets four arguments: 
  + x: sample space in form of a vector
  + size: your desired sample size
  + replace: sampling with/without replacement
  + prob: a vector of probability weights

+ some of the arguments have default values. What are those arguments?

+ How to use (or call) this function? 


---

## Calling a function in R

```{r echo=TRUE}
# Functions arguments can be matched: 1) by position or 2) by name
sampSpace <- 1:6 # rolling a die
sample(sampSpace, 1) # arguments with default values can be omitted
sample(size = 1, x = sampSpace) # no need to remember the order 
sample(size = 1, sampSpace)
```


---

## Writing your Own functions

```{r echo=TRUE, eval=FALSE}
yourFnName <- function(<your arguments>){
  # body of your code
  
  # return the output of the function
}

# to use your function, you can simply call the function name as:
yourFnName(<set values for the input arguments>)
```


---

## Writing your Own functions

+ Let's write a function that gets three arguments: a, b, c
+ The function then returns min of these two numbers

```{r echo=TRUE, eval=TRUE}
myMin <- function(a, b, c){
  myMinVal <- min(a, b, c)
  return(myMinVal)
}

myMin(10, 20, 30)
myMin(10, NA, 20) # ? how to fix this so it returns 10
```


---

## Some notes on Functions

> 1. Variables defined within a function are locally defined (i.e. not defined outside of the function).

> 2. Functions in R are treated like any other first class objects. This means functions can be passed as arguments of other functions.

> 3. Arguments of functions are evaluated as they are needed (lazy evaluation). 

> 4. " ... " can be an argument of a function and it refers to a situation where number of input arguments can be varied and is not fixed upfront. 


---

## Lazy Evaluation of Function Arguments

```{r echo=TRUE, error=T}
myLazyFn1 <- function(a, b){
  return(a)
}
myLazyFn1(10) # No error!

myLazyFn2 <- function(a, b){
  print(a)
  print(b)
  return(1)
}
myLazyFn2(10) 
```


---

## Some useful functions:

+ Here we are going to talk about:
  + str(): a function to explain internal structure of a function
  + apply(): to apply a function to a matrix or dataframe
  + lapply(), sapply(), tapply(), mapply(): applying a function to a vector
  + split(): to split a dataset by levels of a factor


---

### str():

+ str() is a compact way of understanding what an object is and what is in that object.
```{r echo=TRUE}
str(str)
str(sample)

genderF <- factor(sample(c("Male", "Female"), 20, replace = TRUE))
str(genderF)
```


---

### str():
```{r echo=TRUE}
myMat <- matrix(1:10, ncol = 5)
str(myMat)
myList <- list(numVec = 1:3, logVec = F, charVec = LETTERS[1:4])
str(myList)
```


---

### apply():

```{r echo=TRUE}
str(apply) # try ?apply for more info
```

+ apply() is a useful function to apply a function (FUN) on a MARGIN of a matrix or dataframe (X)

+ MARGIN: a vector giving the subscripts which the function will be applied over
  + 1: indicates rows
  + 2: indicates columns
  + c(1, 2): indicates rows and columns

+ FUN: refers to the function that we want to apply on the dataset

+ "..." : additional arguments of FUN


---

### apply():

```{r echo=TRUE}
myMat <- matrix(1:10, ncol = 5)
myMat[2,c(2, 5)] <- NA
myMat
apply(myMat, 2, sum, na.rm = TRUE)
```


---

### apply():

```{r echo=TRUE}
# consider iris dataset: 
head(iris) # more info ?iris
# suppose we are interested in getting 25% and 75% of each numeric column
```


---

### apply():

```{r echo=TRUE}
# Consider iris dataset: 
apply(iris[,-5], 2, quantile, probs = c(0.25, 0.75))
```


---

### lapply() and sapply():

```{r echo=TRUE}
str(lapply)
str(sapply)
```
+ x: a list, dataframe, or a vector
+ FUN: the function to be applied to each element of X
+ "...": other arguments of FUN


---

### lapply() and sapply():
```{r echo=TRUE}
myList <- list(e1 = 1:10, e2 = -1:-10)
lapply(myList, mean)
sapply(myList, mean)
```


---

### lapply() v. sapply()?:

+ sapply() simplifies the result of lapply.
  
+ If the result of lapply is a list with all elements of the same length:
    + if length == 1: sapply() returns a vector
    + if length != 1: sapply() returns a matrix
    
+ otherwise, sapply() generates a list similar to lapply()


---

### lapply() & sapply() with a user-defined FUN
```{r echo=TRUE}
myList <- list(e1 = 1:10, e2 = -1:-10)
lapply(myList, function(element){return(mean(c(element[1], element[length(element)])))})
sapply(myList, function(element){return(mean(c(element[1], element[length(element)])))})
```


---

### tapply():
```{r echo = TRUE}
str(tapply)
```

+ tapply() applies a function on a subset of a vector
+ X: is a vector 
+ INDEX: list of one or more factors, each of same length as X
+ FUN: our function of interest
+ "...": other arguments of FUN
+ simplify: any guess???


---

### tapply():
```{r echo = TRUE}
HeightDF <- data.frame(height = c(rnorm(100, 180, 3), rnorm(100, 170, 3)), gender = factor(rep(c("M", "F"), each = 100)))
head(HeightDF)
tapply(HeightDF$height, HeightDF$gender, mean)
```


---

### mapply():
```{r echo=TRUE}
str(mapply)
```
  
+ all previous "apply" functions were univariate
  + f(x, {some other parameters})
  
+ What to do if we want to apply a multivariate function:
  + f(x, y, {some other parameters}) # we can have more than 2 variables 


---

### mapply():
```{r echo=TRUE}
l1 <- list(e1 = 1:10, e2 = 1:10)
l2 <- list(e1 = -1:-10, e2 = -1:-10)
# how to get l1$e1[i] + l1$e2[i] + l2$e1[i] + l2$e2[i] ? 
mapply(sum, l1$e1, l1$e2, l2$e1, l2$e2)
```


---

### split():
```{r echo=TRUE}
str(split) # ?split for more info
```

+ X: a vector or a data frame
+ f: factor
+ drop: should R drop empty factor levels?


---

### split():
```{r echo=TRUE}
str(HeightDF)
# Goal: to separate Females from Males
splitData <- split(HeightDF$height, HeightDF$gender)
str(splitData)
```


---


## Time for Break for 10 Minutes :)


---


