---
title: "Intro to R Workshop: Session 3"
#author: "Yuxiao Wang"
date: "May 30, 2016"
output: slidy_presentation
subtitle: UCI Data Science Initiative
---

```{r, include=FALSE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

---


## Session 3 - Agenda

1. Useful Matrix Functions
2. Statistical Distributions in R

---


## Useful Matrix Functions

Consider matrix "A". We can then have:

> 1. t(A): transpose of A
> 2. solve(): to get inverse of A
> 3. eigen(): to get eigen values and eigen vectors (if A is symmetric)
> 4. We only cover solve() in this lecture

---

### solve():
+ Consider B = A %*% X (where X is an unknown matrix)
+ Then: X = solve(A, B)
+ In a special case where B = I, X = $A^{-1}$

```{r echo=TRUE}
A <- matrix(c(1, 2, 3, 2, 4, 5, 3, 5, 6), ncol = 3)
A
# to get inverse of A: solve(A)
```

---

### solve():
```{r echo=TRUE}
solve(A)
# To check that solve(A) is inverse of A:
solve(A)%*%A
```
+ Machine epsilon is defined to be the smallest positive number which, when added to 1, gives a number different from 1.
+ Please visit http://en.wikipedia.org/wiki/Machine_epsilon for more info

---

## Statistical Distributions in R:

+ R has many in-built statistical distributions
  + examples: binomial, poisson, normal, chi square, ...

+ Each distribution in R has four functions:
  + these functions begin with a "d", "p", "q", or "r" and are followed by the name of the distribution

+ ddist(parameters): refers to the density of each distribution
+ rdist(parameters): generates random numbers out of each distribution
+ qdist(parameters): to get quantile of a distribution
+ pdist(parameters): to calculate CDF


---

### Example of a Discrete Distribution:
```{r echo=TRUE}  
# Consider tossing a coin 10 times
str(dbinom)
dbinom(5, 10, 0.5) # prob of getting five heads

str(pbinom) # cumulative dist
pbinom(5, 10, 0.5) # Pr[X <= 5]
```


---

### Example of a Discrete Distribution:
```{r echo=TRUE}  
str(qbinom) # quantile: Pr[X <= ?] = known value
qbinom(0.6230, 10, 0.5) # get the value of ? s.t. Pr[X <= ?]=0.6230

str(rbinom) # Generating random numbers
rbinom(20, 10, 0.5) # 20 ind samples from binomial(10, 0.5)
```


---

### Example of a Continuous Distribution:
```{r echo=TRUE}  
# Consider a standard Normal distribution
str(dnorm)
dnorm(x = 0, mean = 0, sd = 1, log = FALSE)

str(pnorm) # cumulative dist
pnorm(0, mean = 0, sd = 1)
```


---

### Example of a Continuous Distribution:
```{r echo=TRUE}  
str(qnorm) # quantile
qnorm(0.5, mean = 0, sd = 1)

str(pnorm) # cumulative dist
rnorm(10, mean = 0, sd = 1)
```


---

### Example of a Continuous Distribution:
```{r echo=TRUE, fig.height = 4.5, fig.align='center'}  
# Let's try plotting Normal curve (more on plotting later)
x <- seq(from = -3, to = 3, by = 0.05)
y <- dnorm(x, mean = 0, sd = 1)
plot(x, y, type = "l")
```

---

## Time for Break


