---
title: 'Intro to R Workshop: Session 2'
output: ioslides_presentation
date: "Nov 18, 2016"
subtitle: UCI Data Science Initiative
---

```{r, include=FALSE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Session 2 - Agenda

1. Vectorized Operations in R
2. Reading and Writing Data in R
3. Control Structures
4. R Packages and Functions

## Vectorized Operations

+ R is capable of vectorized operations without any need for running loops
+ This means that a function applied to a vector is actually applied individually to each element
```{r echo=TRUE}
x <- 1:5
y <- c(1, 2, 6, 7, 10)
x + y # R does an element by element summation
x < y
```

## Vectorized Operations

+ Similar to vectors, vectorized operations can be performed for matrices
```{r echo=TRUE}
x <- matrix(1:9, ncol = 3)
y <- matrix(rep(c(5,6,7), 3), ncol = 3)
x + y # R does an element by element summation
x < y
```


## Reading and Writing Data

**The slides for "Reading and Writing Data" section were mainly from Dr. Roger D. Peng, Associate Professor at Johns Hopkins**

Main functions for reading data into R:

1. ```read.table()```, ```read.csv()```: to read tabular data 
2. ```readLines()```: to read lines of a text file
3. ```source()```, ```dget()```: to read R code
4. ```load()```: to read saved workspaces

+ Only ```read.table()``` and ```read.csv()``` are covered in this lecture. 


## Reading and Writing Data

Main functions for writing data from R:

1. ```write.table()```, ```write.csv()```: to write tabular data to file
2. ```writeLines()```: to write lines to a text file
3. ```dump()```, ```dput()```: to write R code to a file
4. ```save()```: to save a workspace

+ Only ```write.table()``` is covered in this lecture


## read.table():

+ ```read.table()``` is the most commonly used function to read data in R
+ Type ```?read.table``` in your R console to see the important arguments in the function
  

+ ```read.csv()``` is intended for reading comma separated value files 
    + It is equivalent to read.table() with sep = "," and header = TRUE



## read.table():

```{r echo = TRUE, eval=TRUE}
irisFile <- read.table(file = "iris.csv", sep=",", header = TRUE)
head(irisFile)
```





## write.table():

```{r echo = T, eval=FALSE}
write.table(irisFile, file = "new_iris.csv", sep = ",", col.names = TRUE)
```

## Control Structures:

+ A control structure is a block of programming that analyzes variables and chooses a direction in which to go based on given parameters
+ Control structures in R include:
    + **for loops**
    + **if/else statements**
    + while loops 
    + repeat
    + break 
    + next
    + return
  

## for loops:

+ Suppose we want to write "The year is [year]" where year is equal to 2014, 2015, and 2016. 
+ One way to do so is like this:

```{r echo = T}
print(paste("The year is", 2014))
print(paste("The year is", 2015))
print(paste("The year is", 2016))
```


## for loops:

+ Or, we can use a for loop: 

```{r echo = T}
for(i in 2014:2016){
  print(paste("The year is", i))
}
```

## for loops:

+ Suppose we have a numeric vector and we want to square each element
```{r echo = TRUE}
vec <- seq(2, 20, by = 2)
```
+ First, create a new vector of the same length as ```vec```
```{r echo = TRUE}
newvec <- vector("numeric", length = length(vec))
```
+ Then, write the for loop
```{r echo = TRUE}
for(i in 1:length(vec)){
  newvec[i] <- vec[i]^2
}
newvec
```

## if/else statements:

+ if/else statements are used to write conditional statements

```{r echo = T}
x <- 7
if (x < 10){
  print("x is less than 10")
}else{
  print("x is greater than 10")
}
```

## Combining for loops and if/else statements:
+ Suppose we have the ages of 10 individuals, and we want to categorize each age as young, middle-aged, or old
+ Using a for loop, we can iterate through each age and then use if/else statements to classify each age

## for loops and if/else statements:
```{r echo = T}
age <- sample(1:100, 10)
ageCat <- rep(NA, length(age))
for (i in 1:length(age)) {
    if (age[i] <= 35){
       ageCat[i] <- "Young"
      }else if (age[i] <= 55){
        ageCat[i] <- "Middle-Aged"
      }else{
         ageCat[i] <- "Old"
      } 
}
age.df <- data.frame(age = age, ageCat = ageCat)
age.df[1:3,]
```


## Functions and Packages:

1. R has many built-in functions
2. Each function has a name followed by (), e.g., ```mean()```
3. Arguments of a function are put within the parentheses
4. R packages are a way to maintain collections of R functions and data sets
5. Packages allow for easy, transparent and cross-platform extension of the R base system


## Functions and Packages:

Terminology:

1. Package: an extension of the R base system with code, data and documentation in a standardized format
2. Library: a directory containing installed packages
3. Repository: a website providing packages for installation
4. Source: the original version of a package with human-readable text and code
5. Base packages: part of the R source tree, maintained by R Core

+ for more info on how R packages are developed, please read: "Creating R Packages: A Tutorial" (Friedrich Leisch)
  + http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf


## How to install a package in R:

There are two main ways to install a package in R:

1. Installing from CRAN: install a package directly from the repository
    + Using R studio: tools/install packages
    + From R console: ```install.packages()```
2. Installing from Source: first download the add-on R package and then type the following in your console:
    + ```install.packages("path_to_file", repos = NULL, type = "source")``` 

+ Once you install a package, you need to load it into R using the function ```library()```


## Popular Packages in R:

1. To visualize data:
  + ggplot2: to create beautiful graphics
  + googleVis: to use Google Chart tools
  
2. To report results:
  + shiny: to create interactive web-based apps
  + knitr: to combine R codes and Latex/Markdown codes
  + slidify: to build HTML 5 slide shows
  
3. To write high-performance R code:
  + Rcpp: to write R functions that call C++ code
  + data.table: to organize datasets for fast operations
  + parallel: to use parallel processing in R



## Functions in R

+ Consider the function sample(). Run ?sample to read the help file or str(sample) to see its arguments. 
```{r echo=TRUE}
str(sample)
```

+ sample() has four arguments: 
    + x: vector of elements from which to choose
    + size: desired sample size
    + replace: sampling with/without replacement (logical)
    + prob: vector of probability weights

+ The help file will specify which arguments have default values (and what those values are)


## Calling a function in R

Function arguments can either be matched by position within the parentheses or by name

```{r echo=TRUE} 
sampSpace <- 1:6 
sample(sampSpace, 1) # arguments with default values can be omitted
sample(size = 1, x = sampSpace) # no need to remember the order 
sample(size = 1, sampSpace)
```


## Writing Your Own Functions

+ One strength of R is the ability of the user to add functions
+ The structure of a function is as follows: 

```{r echo=TRUE, eval=FALSE}
yourFnName <- function(arg1, arg2, ...){
  statements # body of your code
  
  return(object) # what is to be returned
}
```

+ To use your function, you can simply call the function name as:
```{r echo=TRUE, eval=FALSE}
yourFnName(arg1, arg2, ...)
```

## Writing Your Own Functions

+ Let's write a function that takes three values (arguments) a, b, and c and returns the min of the three numbers

```{r echo=TRUE, eval=TRUE}
myMin <- function(a, b, c){
  myMinVal <- min(a, b, c)
  return(myMinVal)
}

myMin(10, 20, 30)
myMin(10, NA, 20) # how to fix this so it returns 10?
```



## Some Useful Functions:

+ Here we are going to talk about:
    + str(): a function to explain internal structure of an object
    + apply(): to apply a function to a matrix or dataframe


## str():

+ str() is a compact way of understanding what an object is and what is in that object
```{r echo=TRUE}
str(str)
str(sample)

genderF <- factor(sample(c("Male", "Female"), 20, replace = TRUE))
str(genderF)
```




## str():
```{r echo=TRUE}
myMat <- matrix(1:10, ncol = 5)
str(myMat)
myList <- list(numVec = 1:3, logVec = F, charVec = LETTERS[1:4])
str(myList)
```





## apply():

```{r echo=TRUE}
str(apply) # try ?apply for more info
```

+ apply() is a function that applies a function (FUN) on a MARGIN of a matrix or dataframe (X)

+ MARGIN: a vector giving the subscripts which the function will be applied over
    + 1: indicates rows
    + 2: indicates columns
    + c(1, 2): indicates rows and columns

+ FUN: refers to the function that we want to apply on the dataset

+ "..." : additional arguments of FUN


## apply():

+ Suppose we have a matrix and we want to sum each column. We can use apply to do this.

```{r echo=TRUE}
myMat <- matrix(1:10, ncol = 5)
myMat
apply(myMat, 2, sum)
```


## apply()

+ Now, suppose our matrix has some NAs that we want to ignore in our summation. 
+ The sum() function has an argument na.rm that ignores NAs in a sum. We can include this in the apply() function. 

```{r echo=TRUE}
myMat <- matrix(1:10, ncol = 5)
myMat[2,c(2, 5)] <- NA
myMat
apply(myMat, 2, sum, na.rm = TRUE)
```

## apply():

+ Consider the iris dataset: 
```{r echo=TRUE}
head(iris) # more info ?iris
```
+ Suppose we are interested in getting the 25% and 75% quantiles of each numeric column

+ Check the help page of quantile() to see what arguments should be included. 

## apply():



```{r echo=TRUE}
apply(iris[,-5], 2, quantile, probs = c(0.25, 0.75))
```


## Other functions in the apply() family:
+ As mentioned above, apply() applies a function on a matrix or an array and returns a vector or an array
+ Other apply functions can be applied to and/or return other data structures (we will briefly mention them here; check the help files for each function for more information)

## Other functions in the apply() family:

+ lapply(): can be used on a list, data frame, or vector; returns a list
+ sapply(): works like lapply(), but simplifies the result
+ mapply(): stands for "multivariate" apply; applies a function to multiple list or multiple vector arguments
+ tapply(): applies a function on a subset of a vector broken down by a factor



## Break Time





