---
title: 'Intro To R Workshop: Session 1'
subtitle: 'UCI Data Science Initiative'
author: "Emma Smith"
date: January 20, 2017
output: ioslides_presentation
---

## Introduction

+ The class will include 5 sessions: 
    + Session 1  (9-10:20): Data Types
    + Session 2  (10:30-11:20): Control Structures and Functions
    + Session 3  (11:30-12): Statistical Distributions
    + Exercise 1 (12:30-1:20): Basic Data Exploration
    + Session 4  (1:20-2:50): T-Tests and Data Visualization 
    + Session 5  (3:00-4:20): Linear Regression
    + Exercise 2 (4:20-5:00): Data Visualization & Statistical Analysis
  



```{r,echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(ggplot2)
```

## Introduction 

+ Feel free to ask questions anytime during lectures.

+ To access this presentation and the codes used during the workshop please visit:
    + https://github.com/UCIDataScienceInitiative/IntroR_Workshop





## Session 1 - Agenda

1. RStudio
2. Data Types in R
3. Data Structures in R
3. Subsetting in R



## What is R?

+ R is a free software environment for statistical computing and graphics
    + See http://www.r-project.org/ for more info 
+ R compiles and runs on a wide variety of UNIX platforms, Windows and Mac OS
+ R is open-source and free
+ R is fundamentally a command-driven system
+ R is an object-oriented programming language 
    + Everything in R is an object (data, functions, etc.)
+ R is highly extendable
    + You can write your own custom functions
    + There are 8000 free add-on packages

 


## R Studio:

1. RStudio is a free and open source integrated development environment (IDE) for R.
2. To download RStudio please visit: http://rstudio.org/
3. Please note that you must have R already installed before installing R Studio.



## Fundamentals

+ When you type commands at the prompt '>' and hit ENTER
    + R tries to interpret what you've asked it to do (evaluation)
    + If it understands what you've written, it does it (execution)
    + If it doesn't, it will likely give you an error or a warning

+ Some commands trigger R to print to the screen, others don't

+ If you type an incomplete command, R will usually respond by changing the command prompt to the '+' character
    + Hit ESC on a MAC to cancel
    + Type _Ctrl_ + *C* on Windows and Linux to cancel



## Data Types in R:


+ R has 5 main atomic data types:
    + Numeric
    + Integer
    + Complex
    + Logical
    + Character



## Data Structures in R:

1. One-dimensional:
    + Vectors
  
2. Multi-dimensional:
    + Matrices
    + Data frames

&nbsp;

+ Everything in R is an object
+ Objects can have attributes
    + e.g., names, dimension, length


## Vectors in R:

+ A vector is the most basic object in R
+ It is one-dimensional; its single dimension is its length
+ A vector of length *n* has *n* cells
+ Each cell can hold a single value, like a numeric or string value
    + In general, vectors can only hold ONE type of data 
  
```{r echo=TRUE}
numVec <- c(2,3,4)      # <- is the assigning operator
numVec
```



## Examples of Vectors

Examples of character, logical, and complex vectors:

```{r echo=TRUE}
charVec <- c("red", "green", "blue")
charVec

logVec <- c(TRUE, FALSE, FALSE, T, F)
logVec

compVec <- c(1 + 0i, 3 + 1i)
compVec
```







## Special Values:

There are some special values in R:

  + Use L to refer to an integer value, e.g., 1L
  + R knows infinity: Inf, -Inf
  + NaN: refers to "Not a number"
  
```{r echo=TRUE}
intVec <- c(1L, 2L, 3L, 4L) 
intVec

a <- Inf; b <- 0
rslt <- c(b/a, a/a)
rslt
```



## Data Type Coercion:

+ In general, vectors CANNOT have mixed types of objects
    + Exception: lists in R 

```{r echo=TRUE, results='hide'}
numCharVec <- c(3.14, "a")
numCharVec                 # What do you expect to be printed? 

numLogVec <- c(pi, T)
numLogVec                   

charLogVec <- c("a", TRUE)
charLogVec 
```

+ The above are examples of implicit coercion 
+ Explicit coercion is also possible



## Data Type Coercion:

+ as(): explicitly coerces objects from one type to another

```{r echo=TRUE}
numVec <- seq(from = 1200, to = 1300, by = 15)
numVec
numToChar <- as(numVec, "character")
numToChar
logVec <- c(F, T, F, T, T)
as(logVec, "numeric")
```



## Data Type Coercion:

+ Coercion does not always work! Be careful about warnings:

```{r echo=TRUE}
compVec <- c(12+10i, 1+6i, -3-2i)
as(compVec, "numeric")

charVec <- c("2.5", "3", "2.8", "1.5", "zero")
as(charVec, "numeric")
```



## Factors:

+ A factor is a vector object used to specify a discrete classification (categorical values).
+ Factors can be ordered or un-ordered
+ Levels of a factor are better labeled (self-descriptive)
    + Consider gender as (0, 1) as opposed to labeled ("F", "M")
```{r echo=TRUE}
Gender <- rep(c("Female", "Male"), times = 3)
Gender
GenderFac1 <- factor(Gender)
GenderFac1
```



## Factors:

```{r echo=TRUE}
levels(GenderFac1)
table(GenderFac1)
unclass(GenderFac1) # returns the factor as integer values
```



## Factors:
```{r echo=TRUE}
GenderFac1 # levels are ordered alphabetically - 1st level = BaseLevel
GenderFac2 <- factor(Gender, levels = c("Male", "Female"))
GenderFac1
GenderFac2
```



## Missing Values:

+ There are two kinds of missing values in R:
    + NaN: stands for "Not a Number" and is a missing value produced by numerical computation.
    + NA: stands for "Not Available" and is used when a value is missing

+ NaN is also considered as NA (the reverse is NOT true)
```{r echo=TRUE}
a <- c(1,2)
a[3]

b <- 0/0
b
```



## Matrices:

+ A matrix is a special case of a vector
    + Unlike vectors, matrices have a dimension attribute

```{r echo=TRUE}
myMat <- matrix(nrow = 2, ncol = 4)
myMat
attributes(myMat)
```



## Matrices:
```{r echo=TRUE}
myMat <- matrix(1:8, nrow = 2, ncol = 4)
myMat # matrices are filled in column-wise
```



## A matrix is a special case of a vector:

```{r echo=TRUE}
myVec <- 1:8
myVec
dim(myVec) <- c(2,4)
myVec
```

+ Similar to vectors, all elements of a matrix should be of the same data type
    + If not, R automatically coerces



## Other Ways to Create a Matrix:

+ Intuitively, matrices seem to be a combination of vectors that are put next to each other (either column-wise or row-wise).

+ rbind() and cbind() (row bind and column bind) achieve this: 
```{r echo=TRUE}
vec1 <- 1:4
vec2 <- sample(1:100, 4, replace = FALSE)
vec3 <- rnorm(4, mean = 0, sd = 1)
colMat <- cbind(vec1, vec2, vec3)
colMat
```



## Other Ways to Create a Matrix:

```{r echo=TRUE}
vec1 <- 1:4
vec2 <- sample(1:100, 4, replace = FALSE)
vec3 <- rnorm(4, mean = 0, sd = 1)
rowMat <- rbind(vec1, vec2, vec3)
rowMat
```







## Lists:

+ Think of a list as a vector with the following main differences: 
    + Each element of a list can have its own data structure regardless of other elements
    + This means, each element can be of a different data type and a different length
```{r echo=TRUE}
myVec <- c(10, "R", 10-5i, T)
myVec
```



## Lists:

```{r echo=TRUE}
myList <- list(10, "R", 10-5i, T)
myList
```
+ Elements of a list are shown with [[]]
+ Elements of a vector are shown with []



## Data Frames:

+ A data frame is a special list where all objects have equal length
+ A data frame looks very similar to a matrix; however, different columns in a data frame can be different data types
```{r echo=TRUE}
studentID <- paste("S#", sample(c(6473:7392), 10), sep = "")
score <- sample(c(0:100), 10)
gender <- sample(c("female", "male"), 10, replace = TRUE)
data <- data.frame(studentID = studentID, score = score, gender = gender)
head(data)
```



## Subsetting:

+ Sometimes, we want to take a subset of a vector, matrix, list, or data frame
+ Consider three main operators to take a subset of an object
    + [ ] single brackets return an object of the same class of the original object
    + [[ ]] double brackets are used primarily for lists and dataframes
    + $ used primarily for lists and dataframes (similar to double brackets)
        + We use $ when selecting an element by name
+ [ ] allows us to select more than one element 
+ [[ ]] and $ allow us to select only one 
      
    
   



## Subsetting Examples:

```{r echo=TRUE}
myVec <- 1:10
myVec[3]

myList <- list(obj1 = "a", obj2 = 10, obj3 = T, obj4 = 10-5i)
myList[[3]]
myList$obj4
```



## Subsetting with [ ]:

+ By using single brackets, we can choose more than one element of an object
```{r echo=TRUE}
x <- seq(from=0, to=100, by=10) 
x
x[1:3]           # select the first, second, and third elements
x[c(2,4,6)]      # select the second, fourth, and six elements
```






## Subsetting with [ ]: Index Vectors

+ Another way to select more than one element from an object is by using index vectors
    + An index vector is a vector of indices that is used to select a subset of another vector (or matrix)
  
```{r echo=TRUE}
x <- seq(from=0, to=100,by=10) 
x
IndVec <- c(1, 2, 3, 4, 5)       # index vector to select the first 5 elements 
x[IndVec]
```



## Index Vectors:

+ There are four types of index vectors:
    1. Logical index vector
    2. Vector of positive integers
    3. Vector of negative integers
    4. Vector of character strings



## Example:

+ Suppose we have grades of ten students

```{r echo=TRUE}
grades <- sample(0:100, 10)
names(grades) <- sample(letters[1:10], 10)
grades
```

+ We will explore the different ways to subset this vector using index vectors



## 1. Logical Index Vector

+ A vector of TRUE/FALSE values that should be the same length as the vector from which we are subsetting. 
    + Values corresponding to TRUE in the index vector are selected

```{r echo=TRUE}
logIndVec <- rep(c(T, F), each = 5)
logIndVec

grades[logIndVec]
```



## 1. Logical Index Vector

+ Logical index vectors can also be generated by using conditional statements
    + Using ==, !=, <, >, ...

```{r echo=TRUE}
logIndVec <- grades > 60
logIndVec

grades[logIndVec]
```



## 2. Index Vector of Positive Integers
+ A vector of positive integers corresponding to the elements you want to subset
+ All of the values in this type of index vector must lie in 1:(length(x)).
```{r echo=TRUE}
posIndVec <- 4:7
posIndVec

grades[posIndVec]
```




## 3. Index Vector of Negative Integers
+  A vector of negative integers indicates the values to be excluded from the vector
```{r echo=TRUE}
negIndVec <- -1:-5
negIndVec

grades[negIndVec]
```




## 4. Vector of Character Strings
+ If a vector has a name attribute, we can take a subset of the vector by calling the names of the elements
```{r echo=TRUE}
chIndVec <- c("a")
chIndVec

grades[chIndVec]
```




## Subsetting Matrices: 
+ We also use the single square brackets to subset matrices
    + In the square brackets, the first position refers to the row(s) and the second position refers to the column(s)
```{r echo=TRUE}
myMat <- matrix(1:8, ncol = 4)
myMat
```

+ Let's go over the various ways to subset this matrix




## Subsetting Matrices: 

```{r echo=TRUE}
myMat[1,1]            # retrieve the element in the first row, first column
myMat[2,]             # retrieve the second row
myMat[,3]             # retrieve the third column
```  



## Subsetting Matrices:

+ By default, when the retrieved elements of a matrix look like a vector, R drops their dimension attribute. We can turn this feature off by setting drop = FALSE

```{r echo=TRUE}
myMat[1,1]
myMat[1,1, drop = FALSE]

myMat[2,, drop = FALSE]
``` 



## Subsetting Lists:
```{r echo=TRUE}
myList <- list(ID = paste("ID", sample(c(100:199), 3), sep = ""), Age = sample(c(18:99), 3), Gender = sample(c("M", "F"), 3, replace = TRUE))
myList
myList[1] # subset is still a list
``` 



## Subsetting Lists:
```{r echo=TRUE}
myList[1:2] # return the first two objects; subset is still a list
myList[[1]] # return the 1st object; subset is a character vector
myList$ID # alternative to [[]]
``` 



## Subsetting Lists:
```{r echo=TRUE}
myList[[1]][2] # return the 2nd element of the 1st object
myList$ID[2]
myList[[c(1,2)]]
``` 



## Subsetting Data Frames:
```{r echo=TRUE}
library(datasets)
data(quakes) # ?quakes for more info
str(quakes)
``` 

## Subsetting Data Frames:

```{r echo=TRUE}
quakes[1:10,]
``` 


## Subsetting Data Frames: 

```{r echo=TRUE}
head(quakes$long)
head(quakes[,c("lat", "long")])
```
