---
title: "Intro to R Workshop (Session 4)"
subtitle: "UCI Data Science Initiative"
author: "Steven Brownlee (based on Wenliang He and Sepehr Akhavan's material)"
date: "10/7/2016"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Sessions4 Agenda
1. T-Tests

2. Plotting

3. Linear Regression


## T-Test in R
T-test can be used to draw statistical conclusions about parameters of interest in the data

+ Is the mean of this data different from zero (or another number)?

+ Is the mean of this data different from that data?

+ Is the regression slope coefficient different from zero?


T-tests can be categorized into two groups:

+ 1) One-Sample t-test

+ 2) Two-sample t-test


##  One-Sample T-Test (Create Data)
```{r echo=TRUE}
set.seed(123)
oneSampData <- rnorm(100, mean = 0, sd = 1)

mean(oneSampData)
sd(oneSampData)
```


##  One-Sample T-Test (True Mean Equal to 0)
```{r echo=TRUE}
oneSampTest.0 <- t.test(oneSampData) 
oneSampTest.0
```


```{r echo=TRUE}
names(oneSampTest.0) 
oneSampTest.0$statistic
oneSampTest.0$estimate
```  


##  One-Sample T-Test (True Mean Equal to mu)
```{r echo=TRUE}
oneSampTest.mu <- t.test(oneSampData, mu = 0.3)
oneSampTest.mu
```  


##  Two-Sample T-Test
Two sample t-tests are categorized into 3 groups:

  + T-Test with equal variances
  
  + T-Test with un-equal variances
  
  + Paired T-Test: can be also considered as one-sample t-test on deltas.


##  Two-Sample T-Test (Un-equal Variances)
```{r echo = TRUE}
Samp1 <- rnorm(300, mean = 2.5, sd = 1)
Samp2 <- rnorm(500, mean = 3.0, sd = 1) # notice: not the same sample size
t.test(Samp1, Samp2)  # default assump: unequal variances
```


##  Two-Sample T-Test (Equal Variances)
```{r echo = TRUE}
t.test(Samp1, Samp2, var.equal = TRUE)  # default assump: unequal variances
```


##  Two-Sample T-Test (Paired T Test)
```{r echo = TRUE}
t.test(Samp1, Samp2[1:300], paired = TRUE) # must be of the same sample size
```


## Basic Histogram
```{r echo = TRUE}
hist(Samp1)
```


## Relative Frequency Histogram
With line at the mean
```{r echo = TRUE}
hist(Samp1, main = "Density of Samp1", freq = F, col = "grey") 
abline(v = mean(Samp1), col = "red", lwd = 2)
```


## Overlayed Histograms
```{r echo = TRUE}
hist(Samp1, main = "Densities of Samp1 and Samp2", xlab = "x", freq = F, col = "grey") 
abline(v = mean(Samp1), col = "red", lwd = 2)
hist(Samp2, freq = F, add = T)
abline(v = mean(Samp2), col = "blue", lwd = 2)
```


## Overlayed Histograms with a Legend
```{r echo = TRUE}
hist(Samp1, main = "Densities of Samp1 and Samp2", freq = F, col = "grey") 
abline(v = mean(Samp1), col = "red", lwd = 2)
hist(Samp2, freq = F, add = T)
abline(v = mean(Samp2), col = "blue", lwd = 2)
legend("topright", legend = c("Samp1", "Samp2"), fill = c("grey","white"), bty = "n", cex = 1.3)
```


## Linear Regression Data Set Description
Here we use "Prestige" dataset from "car" package.

+ education: Average education of occupational incumbents, years, in 1971.

+ income: Average income of incumbents, dollars, in 1971.

+ women: Percentage of incumbents who are women.

+ prestige: Pineo-Porter prestige score for occupation, from a social survey conducted in the mid-1960s.

+ census: Canadian Census occupational code.

+ type: Type of occupation. A factor with levels 
    + bc: Blue Collar
    + prof: Professional, Managerial, and Technical
    + wc: White Collar


## Load Data
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
# install.packages("car", dependencies=TRUE)
library(car)
data(Prestige) # load the data
str(Prestige)
```


##  Look at the Data
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
summary(Prestige)
```
See 4 NA values for "type".
Let's look at them.


## Examine Observations with NA values
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
Prestige[is.na(Prestige$type),]
```

Let's recode "newsboys", "babysitters", and "farmers" as blue collar and exclude athletes.
First find their row indexes, and corresponding names.
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
which(is.na(Prestige$type))  # gives index numbers of the NAs in the vector
rownames(Prestige)[which(is.na(Prestige$type))]             # find the row names at those indexes
```


## Recode/Drop NA Values
Recode "newsboys", "babysitters", and "farmers" to blue collar.

Use "rep" to create a vector with 3 elements of "bc"
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
Prestige[c(53, 63, 67),"type"] <- rep("bc", 3)
summary(Prestige$type)
```

Exclude athletes
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
Prestige <- na.omit(Prestige) 
summary(Prestige$type)
```


## Plotting - Histograms
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
hist(Prestige$prestige, col = "grey", 
     main = "Histogram of Prestige Score", xlab = "Prestige Score")
```


## Plotting - Basic Scatter Plot
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
plot(Prestige$education, Prestige$prestige,
     main = "Prestige Score by Education",
     xlab = "Ave. Years of Education", ylab = "Prestige Score")
```


## Plotting - Including Regression/Smoother Lines
abline adds a line to the plot
lm() is a list object that contains regression coefficients
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
plot(Prestige$education, Prestige$prestige,
     main = "Prestige Score by Education",
     xlab = "Ave. Years of Education", ylab = "Prestige Score")
abline(reg = lm(prestige ~ education, data = Prestige), col = "red", lwd = 2)
lines(lowess(Prestige$education, Prestige$prestige), col = "blue", lty = 2, lwd = 2)
legend("topleft",legend = c("Regression Line", "Smoother"), col = c("red","blue"),
       lwd = c(2,2), lty = c(1,2), bty = "n")
```


## Plotting - Scatter Plot Matrix
Use the scatterplotMatrix function from the "car" package
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
# Use direct ordering of the varaibles to control how they are plotted
scatterplotMatrix(Prestige[,c("prestige","education","income","women","census")])
```


## Plotting - Boxplots 
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
boxplot(prestige ~ type, data = Prestige, col = "grey",
        main = "Distribution of Prestige Score by Types",
        xlab = "Occupation Types", ylab = "Prestige Score")
```


## Linear Regression - Fit the Model
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
myReg <- lm(prestige ~ education + income + women, data = Prestige)
myReg
names(myReg)
```


##  Linear Regression - Summary of Fit
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
summary(myReg)
```


##  Linear Regression - "summary" Contents
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
sum.myReg = summary(myReg)
names(sum.myReg) # show different contents

names(myReg) # this is what we had previously
```


##  Linear Regression - Confidence Interval
+ 95% confidence interval for coefficient of 'income'
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
confint(myReg, 'income', level=0.95)
```

+ 95% confidence interval for each coefficient
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
confint(myReg, level=0.95)
```


##  Linear Regression - Adding Variables
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
mod = update(myReg, ~ . + type); summary(mod)
```


##  Linear Regression - Relevel a Factor
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
levels(Prestige$type)
Prestige$type = relevel(Prestige$type, "prof")

levels(Prestige$type)
```


##  Linear Regression - Relevel a Factor
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
mod = update(myReg, ~ . + type); summary(mod)
```


##  Linear Regression - Diagnostics
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE, fig.height=6.5, fig.width=8}
par(mfrow = c(2, 2), oma = c(0, 0, 2, 0))
plot(myReg)
```


##  Linear Regression - Predict
+ Predict the output for a new input
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
newData = data.frame(education=13.2, income=12000, women=12)
predict(myReg, newData, interval="predict")
```


## End of Session 4
BREAK!